---
title: "Variant Calling Analysis Report"
author: "Raissa, John, Rennatha, Liya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cerulean
---
```{r setup, include=FALSE}
# Setup chunk: Load libraries, set options, install packages if needed
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Uncomment this line if packages are not installed
# install.packages(c("openxlsx", "dplyr", "janitor", "stringi", "stringr", "ggplot2", "data.table"))

library(openxlsx)
library(dplyr)
library(janitor)
library(stringi)
library(stringr)
library(ggplot2)
library(data.table)
```


## Data Loading and Exploration

```{r load-data}
# Load the high-quality variants data
df <- fread("hq_allvariants.tsv")

# Explore the data
dim(df)
colnames(df)
unique(df$SAMPLE_ID)
```

## Summary of High-Quality Variants per Sample

```{r summarize-hq}
# Calculate number of HQ variants per sample
number_hq_per_sample <- df %>%
  clean_names() %>%
  group_by(sample_id) %>%
  summarise(n_hq = n(), .groups = "drop")

number_hq_per_sample
```

## Variant Type Classification and Proportions

```{r classify-variants}
# Annotate each variant as Transition or Transversion
variants_type <- df %>%
  clean_names() %>%
  mutate(
    substitution = paste0(ref, ">", alt),
    variant_type = case_when(
      (ref == "A" & alt == "G") | (ref == "G" & alt == "A") ~ "Transition",
      (ref == "C" & alt == "T") | (ref == "T" & alt == "C") ~ "Transition",
      (ref %in% c("A", "T", "C", "G") & alt %in% c("A", "T", "C", "G")) ~ "Transversion",
      TRUE ~ "Other"
    )
  )

# Proportions of variant types per sample
prop_variants_type <- variants_type %>%
  group_by(sample_id, variant_type) %>%
  summarise(count = n(), .groups = "drop_last") %>%
  mutate(prop = count / sum(count) * 100) %>%
  ungroup()

prop_variants_type
```

## Most Common Variant Substitutions

```{r common-substitutions}
# Count of each substitution per variant type
common_variants <- variants_type %>%
  count(variant_type, substitution, sort = TRUE)

common_variants
```

```{r unique-substitutions}
# List unique substitutions
unique_variants <- variants_type %>%
  select(substitution) %>%
  distinct()

unique_variants
```

## Visualization

### High-Quality Variants per Sample

```{r plot-numbers, fig.height=5}
plot_numbers <- ggplot(number_hq_per_sample, aes(x = sample_id, y = n_hq, fill = sample_id)) +
  geom_bar(stat = "identity") +
  xlab("Sample ID") +
  ylab("Number of High-Quality Variants") +
  ggtitle("HQ Variants per Sample") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  guides(fill = FALSE)

plot_numbers
```

### Proportion of Transitions vs. Transversions per Sample

```{r plot-proportions, fig.height=5}
plot_proportions <- ggplot(prop_variants_type, aes(x = sample_id, y = prop, fill = variant_type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(prop, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3, color = "black") +
  xlab("Sample ID") +
  ylab("Proportion (%)") +
  ggtitle("Transitions vs Transversions per Sample") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Transition" = "darkblue", "Transversion" = "orange")) +
  guides(fill = guide_legend(title = "Variant Type"))

plot_proportions
```

### Substitution Counts by Variant Type

```{r plot-substitutions, fig.height=6}
plot_substitutions <- ggplot(common_variants, aes(x = reorder(substitution, n), y = n, fill = variant_type)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Substitution") +
  ylab("Count") +
  ggtitle("Counts of Substitutions by Variant Type") +
  theme_minimal() +
  scale_fill_manual(values = c("Transition" = "skyblue", "Transversion" = "orange")) +
  guides(fill = guide_legend(title = "Variant Type"))

plot_substitutions
```

---

## Conclusion

This report summarizes variant calling results across multiple samples, showing:
- The number of high-quality variants per sample
- The proportions of transitions vs. transversions
- The most common variant substitutions across all data